---
title: "Functions we have learned"
author: "Anita"
date: "10/31/2019"
output: html_document
---

## Class 1
c()
data.frame()
class()
type()
as.character()
as.numeric()
length()
subset() <- same as filter() from tidyverse
round()
unique()
mean()

 
Indexing:
  df[row index, column index] <- use 'Roman Catholics' heuristic (Roman - rows come first, Catholics - columns come second)

Logical operators:
  Comparisons:
    >
    <
    >=
    <=
    ==
    !=
    
  Boolean operators:
    |
    &
  

    

## Class 2

### Tidyverse
filter()
    Example: filter by any of specified values: columnname %in% c()
    filter(df, name %in% c("Pernille", "Gustav"))

arrange()

desc() - descending order
  used in arrange(dataframe, desc(variable))

select() - decreasing size of subset; can be used for ordering
  Examples:
    #remove first column
    df <- select(df,-1)
    
    #remove several columns:
    notouch_sound_taste <- select(df,-c(touch_floor,touch_hands, sound_level_pref:taste_blind))
    
    #rearrange
    rearranged <- select(df,gender,shoesize, everything())

mutate()
  mutated <- mutate(df, breath_hold_min = breath_hold/60)
  df <- mutate(df,cealing = (romberg_open == 120))

%/% returns the whole number of times the number is dividable by something
  e.g. 65 %/% 60 returns 1
       121 %/% 60 returns 2

summarise()
  This function collapses the whole dataframe into a single summary. For this function to work, it has to follow certain pattern.  First, you specify the data frame you want to summarise and then you say what values you want to have in your summary.

See examples in the chunk below, try to run it and make sense of results:

```{r}
#Make a summary with just one value - the average shoesize 
summarise(df,mean(shoesize))

#summary with several values: the average shoesize and its standard deviation
summarise(df,mean(shoesize),sd(shoesize))
```



summarise() is quite useless by itself, but everything changes when we **group** our data!

group_by() 
  group_by() takes an existing data frame and converts it into a data frame grouped by some principle. See examples in the chunk below:

```{r}
#group data by gender
grouped_bygender <- group_by(df, gender)

#group data by native language
grouped_bylanguage <- group_by(df, native_Danish)

```


%>% 
  pipes, reads as: send the resulting dataframe to the next function

## Class 3
### R markdown:
Command 1: find default chunk options (opts_chunk), and set the setting echo (whether to show your code) to TRUE 
knitr::opts_chunk$set(warning=TRUE, message=TRUE)

Command 2: set a new working directory to ALL chunks - not just the current chunk
knitr::opts_knit$set(root.dir = 'relative_path_to_root_from_Rmd' )

You can make a new R code chunk using keyboard shortcut
Ctrl + Alt + I (Cmd + Option + I on macOS). 


## Class 4

## Class 5

## Class 6
t.test(Continuous outcome variable ~ Categorical predictor varibale, data = dataFrame, paired = FALSE/TRUE, var.equal = FALSE/TRUE)

Going through checking normality assumptions and transformation again!

In case if data can't be helped:
WRS2::yuen(Reaction_Time ~ Gender, data = data) for independent t-test
WRS2::yuend(x, y, tr = 0.2) for paired t-test

## Class 7

## Class 8

##